Anomaly detection with kernel density estimation on manifolds

### 2022-07-05 paper/paper.Rmd
Rewrite kde for riem manifolds

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Kernel density estimation on manifolds {#vkde}

@@

## Multivariate kernel density estimation

@@

For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
More importantly, the estimator in Equation \@ref(eq:vkde) is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the underlying density is supported on a single manifold. The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point. This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.

@@

Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$. Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with density $f$. For each point $p\in M$, @Pelletier2005-vu proposed a kernel density estimator of $f$ to be 
\begin{equation}
\label{eq:denriem}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i)}{h}),
\end{equation} 
where $h$ is the global bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q, $\exp_q^{-1}(p)$. The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as 
\begin{equation}
\label{eq:denconsist}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
\end{equation} 
and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].

### Volume density function

@@

@@

## Volume density function

>>>>

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embe
dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Kernel density estimation on manifold learning embeddings {#vkde}

@@

## Multivariate kernel density estimation {#mvkde}

@@

For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
More importantly, the estimator in Equation \@ref(eq:vkde) is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the manifold has a differentiable structure, which has to be taken into account when considering the neighborhood of a sample point. Therefore, we need to consider the probability density estimation on a Riemannian manifold and adapts the techiniques on Euclidean samples to a non-Euclidean setting. This is discussed in \autoref{riemest}.

@@

Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the geodesic distance induced by the Riemannian metric $g$ and the strictly positive injectivity radius of the manifold $inj_g(M)$ [@Chavel2006-mp]. Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with density $f$. To estimate the density of the manifold, @Pelletier2005-vu proposed a kernel density estimator for $f$ at point $\pmb{p}\in M$ as
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
\end{equation} 
where $r$ is the global bandwidth satisfying $r \leq r_0$ for fixed $r_0$ such that $0\leq r_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{\pmb{X}_i}(\pmb{p})$ is the volume density function on the manifold for point $\pmb{X}_i$. The estimator in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case in Equation \@ref(eq:kde) and converges at the same rate as the Euclidean estimator [@Henry2009-ll].

Denote $\lambda$ as the Lebesque measure of $\mathbb{R}^d$. The kernels $K$ in \@ref(eq:denriem) are functions of the geodesic distance of the manifold $d_g$ and they satisfy the following conditions [@Pelletier2005-vu]:
\begin{align}
& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(x)=K(0).
\end{align}
Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, $r\leq inj_g(M)$. Under such condition, for each $\pmb{p} \in M$, there exists a normal coordinate neighborhood of $\pmb{p}$ containing an open ball $B(\pmb{p}, r) \in T_pM$ centered at $\pmb{p}$ with radius $r$.

Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denotes a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$. Then $\frac{Dw_1}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$. 
The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$

For the exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
$$
in normal coordinates. The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.

<!-- $$ -->
<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
<!-- $$ --> 
<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->

# Distotion-corrected kernel density estimator (DC-KDE) {#dckde}

@@

<!-- The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point. This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}. -->

@@

## Volume density function [MOVE TO SECTION 2.2.1]
